### 2020-08-03 paper/sex_selection_spacing-ver4.tex
Fix percentile examples and edit Empirical Strategy section

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that Hindu women's birth intervals
increased over the last four decades, more so the higher the parity and
the more educated the woman. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
However, the likelihood of very short spacing did not change substantially. 
As a result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. 
Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. 
Furthermore, cohort fertility is still at or above replacement level for all but the
best-educated urban women. 
Finally, infant mortality risk has declined substantially over time for all groups, but 
fastest for the lower education groups, who are now close to the level of women with the most
education. 
Short birth spacing is still associated with higher mortality, although substantially less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with
higher infant mortality risk.
\noindent JEL: J1, O12, I1
\noindent Keywords: India, prenatal sex determination, censoring, competing risk, non-proportional hazard
\end{abstract}

@@

Second, even without sex selection, it is unlikely that characteristics, such as 
the sex composition of previous births, have the same effects throughout the entire 
birth interval. 
The proportional hazard model requires that the hazard for any individual is a 
fixed proportion of the hazard for any other individual. 
That the effects are not constant violates that assumption, and the results from a 
proportional hazard model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals since there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

@@

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, splitting spells into smaller intervals---here equal to three months---and 
treating them as observations, I can estimate the model using a standard multinomial 
Logit model. 
In the reorganized data, the outcome variable is 0 if the woman does not have a child 
in a given interval (the base outcome), 1 if she gives birth to a son in that interval, 
and 2 if she gives birth to a daughter in that interval.

@@

For birth intervals, I estimate a set of percentile birth intervals.
To find a given percentile birth interval, I first calculate for each woman when there is 
a given percent chance that she will have given birth, conditional on the probability of 
giving birth in the spell.
I then add nine months to account for the start of the spell. 
For example, with an 80 percent chance of giving birth by the end of the spell, the median birth 
interval is the predicted number of months before a woman passes the 40 percent mark on her 
survival curve. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
parity progression probabilities as weights.
For an individual woman, the estimated percent of births that are boys at $t$, $\hat{Y_t}$, is:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100,
\label{eq:probability_son}
\end{equation}
where $b$ and $g$ are the predicted probabilities of having a boy and a girl, respectively,
conditional on the explanatory variables and not having had a child before $t$.
Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
$t$ gives the predicted percent of boys over the entire spell for an individual.%
\footnote{
Imagine $T=2$. 
If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
}

@@

>>>>

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that Hindu women's birth intervals
increased over the last four decades, more so the higher the parity and
the more educated the woman. 
Women most likely to use sex selection---well-educated with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
However, the likelihood of very short spacing did not change substantially. 
As a result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. 
Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. 
Furthermore, cohort fertility is still at or above replacement level for all but the
best-educated urban women. 
Finally, infant mortality risk has declined substantially over time for all groups, but 
fastest for the lower education groups, who are now close to the level of women with the most
education. 
Short birth spacing is still associated with higher mortality, although substantially less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with
higher infant mortality risk.

\noindent JEL: J1, O12, I1

\noindent Keywords: India, prenatal sex determination, censoring, competing risk, 
non-proportional hazard
\end{abstract}

@@

Second, even without sex selection, it is unlikely that characteristics, such as 
the sex composition of previous births, have the same effects throughout the entire 
birth interval. 
The proportional hazard model requires that the hazard for any individual is a 
fixed proportion of the hazard for any other individual. 
Non-constant effects violate that assumption, and the results from a proportional hazard 
model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals since there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.

@@

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
The two exit states are the birth of a boy, $j=1$, or a girl, $j=2$.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, splitting spells into smaller intervals---here equal to three months---and 
treating them as observations, I can estimate the model using a standard multinomial 
Logit model. 
% In the reorganized data, the outcome variable is 0 if the woman does not have a child 
% in a given interval (the base outcome), 1 if she gives birth to a son in that interval, 
% and 2 if she gives birth to a daughter in that interval.

@@

For birth intervals, I estimate a set of percentile birth intervals. 
I first calculate for each woman when there is a given percent chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80 percent parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60 percent mark on her survival 
curve. 
I then and add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

% The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
% parity progression probabilities as weights.
% For an individual woman, the estimated percent of births that are boys at $t$, $\hat{Y_t}$, is:  
% \begin{equation}
% \hat{Y}_t 
% = 
% \frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
% { P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
% \times 100,
% \label{eq:probability_son}
% \end{equation}
% where $b$ and $g$ are the predicted probabilities of having a boy and a girl, respectively,
% conditional on the explanatory variables and not having had a child before $t$.
% Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
% $t$ gives the predicted percent of boys over the entire spell for an individual.%
% \footnote{
% Imagine $T=2$. 
% If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
% then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
% }

The predicted sex ratio is the weighted average of individual predicted sex ratios, using 
parity progression probabilities as weights.
To find the individual sex ratio, I estimate the percent of births that are boys at $t$, 
conditional on not having had a child before $t$.
Weighting the percentages boys with the likelihood of exiting the spell with a birth across all 
$t$ gives the predicted percent of boys over the entire spell for an individual.%
\footnote{
Imagine $T=2$. 
If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
}

@@

\section{Empirical Model Details}

[More formal version of empirical strategy section]

The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---the model can be estimated using a standard multinomial logit model.

The distribution of spacing is captured by the survival curve, which shows the probability 
of not having had a birth yet by spell duration, for a given set of explanatory variables.
The survival curve at time $t$ is 
\begin{equation}
\label{eq:survival}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

% Because the probability of ever having a next birth varies across groups, a direct 
% comparison of standard survival curves tells us little about how the spread of sex 
% selection affects birth spacing across groups.
% I, therefore, condition on the predicted likelihood of parity progression when examining 
% birth spacing measures, such as the average duration to a birth.
% The reliability of this approach depends on whether the spell length covered is 
% sufficiently long that few women are likely to give birth after the spell cut-off.
% I discuss the choice of spell length below.%
% \footnote{
% It is important to note that the approach is not the same as merely calculating 
% the birth spacing measures for women who already have a given 
% parity child in the survey because that number does not take into account
% the censoring of spells that will eventually lead to a birth.
% }
% In addition, I present graphs of survival curves conditional on parity progression,
% which therefore begin at 100\% and end at 0\%.

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period, as
\begin{align}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
\label{eq:probability_boy} \\
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
\label{eq:probability_girl}
\end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.%
\footnote{
Imagine $T=2$. 
If 54 and 66 percent of births are boys and the likelihood of giving birth 20 and 40 percent, 
then the predicted sex ratio is $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$ percent boys. 
}
